---
title: "Compare Shrub Cover Data Sources"
author: "Erica Christensen"
date: "4/25/2022"
output: pdf_document
---

```{r setup, include=FALSE}
library(dplyr)
library(terra)
library(ggplot2)
library(lubridate)

# load data

# jornada outline
jer_boundary = vect('shapefile/jer_boundary.shp')

# list RCMAP shrub files
file_list = list.files('RCMAP', pattern = 'rcmap_shrub_[[:alnum:]]*_[[:alnum:]]*.tiff$', full.names=T)

# load rcmap raster
rcmap_2016 = terra::rast(file_list[1])

# convert jer_boundary to same crs as raster
jer_boundary_utm = project(jer_boundary, crs(rcmap_2016))

# crop raster to jornada bounds
rcmap_2016_crop = crop(rcmap_2016, jer_boundary_utm)

# values of 101 in RCMAP seem to be NAs
rcmap_2016_crop[rcmap_2016_crop>100] <- NA



# # ====== do once
# # list RAP shrub cover files
# rap_files = list.files('RAP/shrub rasters MLRA42', pattern = '*.tif', full.names = T)
# # load rap raster and reproject
# rap_2016 = terra::rast(rap_files[32]) %>% project(crs(rcmap_2016))
# # crop
# rap_2016_crop = crop(rap_2016, rcmap_2016_crop)
# 
# terra::writeRaster(rap_2016_crop, filename='RAP/for comparisons/rap_2016_jornadacrop.tif', overwrite=T)
# 
# # load rap raster and reproject
# rap_2017 = terra::rast(rap_files[33]) %>% project(crs(rcmap_2016))
# # crop
# rap_2017_crop = crop(rap_2017, rcmap_2017_crop)
# 
# terra::writeRaster(rap_2017_crop, filename='RAP/for comparisons/rap_2017_jornadacrop.tif', overwrite=T)
# # =========



# load rap raster
rap_2016_crop = terra::rast('RAP/for comparisons/rap_2016_jornadacrop.tif')

# the extents do not match, probably because of different initical crs, so I resample rap
rap_2016_resamp = resample(rap_2016_crop, rcmap_2016_crop)

```

## Compare RAP to RCMAP

RAP and RCMAP are two data products that estimate fractional vegetation cover for the SW United States. In both cases, models are based on LANDSAT, so both have 30m pixel resolution and span 1985-present. 

Looking at the distribution of values in RAP and RCMAP, it's apparent that RAP values are generally higher. [2016 data shown]


```{r rapvsrcmap, echo=F}
rcmapdifference = rcmap_2016_crop - rap_2016_resamp


# look at histograms of rcmap and rap separately
par(mfrow=c(1,3))
hist(values(rcmap_2016_crop), breaks=0:110, main='RCMAP values')
hist(values(rap_2016_resamp), breaks=0:110, main='RAP values')
suppressWarnings(hist(rcmapdifference, breaks= -52:30, main='RCMAP - RAP values'))
```

## Year to year change in shrub

Shrub cover shouldn't change more than about .25% per year. I looked at the difference between 2016 and 2017 for RAP and RCMAP shrub cover estimates. 

```{r pixelwisechange, echo=FALSE}
# look at average change per pixel between 2016 and 2017

# load rcmap raster
rcmap_2017 = terra::rast(file_list[2])

# crop raster to jornada bounds
rcmap_2017_crop = crop(rcmap_2017, jer_boundary_utm)

# values of 101 in RCMAP seem to be NAs
rcmap_2017_crop[rcmap_2017_crop>100] <- NA

# load rap raster
rap_2017_crop = terra::rast('RAP/for comparisons/rap_2017_jornadacrop.tif')

# the extents do not match, probably because of different initical crs, so I resample rap
rap_2017_resamp = resample(rap_2017_crop, rcmap_2017_crop)

# difference in RAP, RCMAP 2016-2017
rap_1yrdiff = rap_2017_resamp-rap_2016_resamp
rcmap_1yrdiff = rcmap_2017_crop-rcmap_2016_crop

# plots
par(mfrow=c(2,2))

hist(values(rap_1yrdiff), breaks=-25:22, main='RAP 1-yr difference', xlab='')
plot(rap_1yrdiff)
hist(values(rcmap_1yrdiff), breaks=-25:22, main='RCMAP 1-yr difference', xlab='')
plot(rcmap_1yrdiff)

```

RCMAP has much lower year to year shrub change than RAP.

## AIM shrub cover data

There are point shrub cover measurements available from AIM TerrADat. I compared point measurements to RAP (top row of figure) and RCMAP (bottom row of figure) where they overlap. Root Mean Squared Error varied by year.

```{r aimdata, echo=F}
# raw data
aim_raw = read.csv('Ground data/BLM_Natl_AIM_TerrADat_Hub.csv')

# get only New Mexico, below 35.4 latitude, get only variables related to date, location, and cover
aim <- aim_raw %>%
  dplyr::filter(State=='NM', Latitude_NAD83<35.4) %>%
  dplyr::select(OBJECTID, PrimaryKey, PlotKey, PlotID, 
                lat = Latitude_NAD83, 
                long = Longitude_NAD83, DateVisited,
                shrubcover = AH_ShrubCover) %>%
  mutate(date = as.Date(DateVisited),
         year = year(date))

# look at one year at a time: 2020
aim2020 = dplyr::filter(aim, year==2020)
aim2020$ID = 1:nrow(aim2020)

# turn into terra SpatVector object
df2020 = data.frame(ID=aim2020$ID, shrubcover=aim2020$shrubcover)
pts2020 = terra::vect(cbind(aim2020$lon, aim2020$lat), atts = df2020, crs="+proj=longlat +datum=NAD83")

# get RAP data for 2020
rap2020 = terra::rast('RAP/shrub rasters MLRA42/RAP_shrubtree_2020.tif')

# get intersection of AIM points and RAP raster
aim_rap_2020 = terra::extract(rap2020, pts2020) %>%
  merge(aim2020) %>%
  dplyr::filter(!is.na(RAP_shrubtree_2020))
rmse2020 = round(sqrt(mean((aim_rap_2020$shrubcover-aim_rap_2020$RAP_shrubtree_2020)^2)), digits=2)

# plot AIM shrub cover vs RAP shrub cover
rap2020plot = ggplot(aim_rap_2020) +
  geom_point(aes(x=shrubcover, y=RAP_shrubtree_2020)) +
  xlab('AIM') +
  ylab('RAP') +
  ggtitle('2020') +
  geom_abline(intercept = 0, slope = 1) +
  coord_cartesian(xlim=c(0,50), ylim=c(0,50)) +
  annotate(geom='text', x=5, y=45, label=paste('RMSE: ',rmse2020)) +
  theme_bw()

# look at one year at a time: 2018
aim2018 = dplyr::filter(aim, year==2018)
aim2018$ID = 1:nrow(aim2018)

# turn into terra SpatVector object
df2018 = data.frame(ID=aim2018$ID, shrubcover=aim2018$shrubcover)
pts2018 = terra::vect(cbind(aim2018$lon, aim2018$lat), atts = df2018, crs="+proj=longlat +datum=NAD83")

# get RAP data for 2018
rap2018 = terra::rast('RAP/shrub rasters MLRA42/RAP_shrubtree_2018.tif')

# get intersection of AIM points and RAP raster
aim_rap_2018 = terra::extract(rap2018, pts2018) %>%
  merge(aim2018) %>%
  dplyr::filter(!is.na(RAP_shrubtree_2018))
rmse2018 = round(sqrt(mean((aim_rap_2018$shrubcover-aim_rap_2018$RAP_shrubtree_2018)^2)), digits=2)

# plot AIM shrub cover vs RAP shrub cover
rap2018plot = ggplot(aim_rap_2018) +
  geom_point(aes(x=shrubcover, y=RAP_shrubtree_2018)) +
  xlab('AIM') +
  ylab('RAP') +
  ggtitle('2018') +
  geom_abline(intercept = 0, slope = 1) +
  coord_cartesian(xlim=c(0,50), ylim=c(0,50)) +
  annotate(geom='text', x=5, y=45, label=paste('RMSE: ',rmse2018)) +
  theme_bw()

# look at one year at a time: 2016
aim2016 = dplyr::filter(aim, year==2016)
aim2016$ID = 1:nrow(aim2016)

# turn into terra SpatVector object
df2016 = data.frame(ID=aim2016$ID, shrubcover=aim2016$shrubcover)
pts2016 = terra::vect(cbind(aim2016$lon, aim2016$lat), atts = df2016, crs="+proj=longlat +datum=NAD83")

# get RAP data for 2016
rap2016 = terra::rast('RAP/shrub rasters MLRA42/RAP_shrubtree_2016.tif')

# get intersection of AIM points and RAP raster
aim_rap_2016 = terra::extract(rap2016, pts2016) %>%
  merge(aim2016) %>%
  dplyr::filter(!is.na(RAP_shrubtree_2016))
rmse2016 = round(sqrt(mean((aim_rap_2016$shrubcover-aim_rap_2016$RAP_shrubtree_2016)^2)), digits=2)

# plot AIM shrub cover vs RAP shrub cover
rap2016plot = ggplot(aim_rap_2016) +
  geom_point(aes(x=shrubcover, y=RAP_shrubtree_2016)) +
  xlab('AIM') +
  ylab('RAP') +
  ggtitle('2016') +
  geom_abline(intercept = 0, slope = 1) +
  coord_cartesian(xlim=c(0,50), ylim=c(0,50)) +
  annotate(geom='text', x=5, y=45, label=paste('RMSE: ',rmse2016)) +
  theme_bw()

# ===================
# RCMAP comparison to AIM
# get RCMAP data for 2020
#rcmap2020 = terra::rast('RCMAP/rcmap_shrub_2020_LajMUGu45oF76trHI4pc.tiff')
rcmap2020 = terra::rast(file_list[5])
rcmap2020[rcmap2020>100] <- NA

# get RCMAP into same crs as points
rcmap2020_lonlat = project(rcmap2020, "+proj=longlat +datum=NAD83")

# get intersection of AIM points and RCMAP raster
aim_rcmap_2020 = terra::extract(rcmap2020_lonlat, pts2020, fun=mean) %>%
  merge(aim2020) %>%
  dplyr::filter(!is.na(Red))
rmse = round(sqrt(mean((aim_rcmap_2020$shrubcover-aim_rcmap_2020$Red)^2)), digits=2)

rcmap2020plot = ggplot(aim_rcmap_2020) +
  geom_point(aes(x=shrubcover, y=Red)) +
  xlab('AIM') +
  ylab('RCMAP') +
  ggtitle('2020') +
  geom_abline(intercept = 0, slope = 1) +
  coord_cartesian(xlim=c(0,50), ylim=c(0,50)) +
  annotate(geom='text', x=5, y=45, label=paste('RMSE: ',rmse)) +
  theme_bw()

# get RCMAP data for 2018
rcmap2018 = terra::rast(file_list[3])
rcmap2018[rcmap2018>100] <- NA

# get RCMAP into same crs as points
rcmap2018_lonlat = project(rcmap2018, "+proj=longlat +datum=NAD83")

# get intersection of AIM points and RCMAP raster
aim_rcmap_2018 = terra::extract(rcmap2018_lonlat, pts2018, fun=mean) %>%
  merge(aim2018) %>%
  dplyr::filter(!is.na(Red))
rmse18 = round(sqrt(mean((aim_rcmap_2018$shrubcover-aim_rcmap_2018$Red)^2)), digits=2)

rcmap2018plot = ggplot(aim_rcmap_2018) +
  geom_point(aes(x=shrubcover, y=Red)) +
  xlab('AIM') +
  ylab('RCMAP') +
  ggtitle('2018') +
  geom_abline(intercept = 0, slope = 1) +
  coord_cartesian(xlim=c(0,50), ylim=c(0,50)) +
  annotate(geom='text', x=5, y=45, label=paste('RMSE: ',rmse18)) +
  theme_bw()

# get RCMAP data for 2016
rcmap2016 = terra::rast(file_list[1])
rcmap2016[rcmap2016>100] <- NA

# get RCMAP into same crs as points
rcmap2016_lonlat = project(rcmap2016, "+proj=longlat +datum=NAD83")

# get intersection of AIM points and RCMAP raster
aim_rcmap_2016 = terra::extract(rcmap2016_lonlat, pts2016, fun=mean) %>%
  merge(aim2016) %>%
  dplyr::filter(!is.na(Red))
rmse16 = round(sqrt(mean((aim_rcmap_2016$shrubcover-aim_rcmap_2016$Red)^2)), digits=2)

rcmap2016plot = ggplot(aim_rcmap_2016) +
  geom_point(aes(x=shrubcover, y=Red)) +
  xlab('AIM') +
  ylab('RCMAP') +
  ggtitle('2016') +
  geom_abline(intercept = 0, slope = 1) +
  coord_cartesian(xlim=c(0,50), ylim=c(0,50)) +
  annotate(geom='text', x=5, y=45, label=paste('RMSE: ',rmse16)) +
  theme_bw()

multifig = gridExtra::grid.arrange(rap2016plot, rap2018plot, rap2020plot, rcmap2016plot, rcmap2018plot, rcmap2020plot, nrow=2)
ggsave('RAP_RCMAP_RMSE.png', plot=multifig)
```